---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setOptions, echo=TRUE}
## loading used libs and setting code to globaly appear along the document
rm(list = ls())
library(ggplot2)
library(knitr)
opts_chunk$set(echo = TRUE)

# setting decimal number output
knit_hooks$set(inline = function(x) {
  prettyNum(round(x,2), big.mark=",")
})

```

## Loading and preprocessing the data
```{r loadingData, echo=TRUE}

if (!("activity.csv" %in% dir())){
  unzip("activity.zip")
}
actData <- read.csv("activity.csv",
                    header = TRUE,
                    colClasses = c("numeric", "Date", "numeric"),
                    na.strings = "NA")

```

## What is mean total number of steps taken per day?

```{r stepsPerDay}

stepsByDay <- aggregate(actData$steps, by = list(actData$date), FUN = sum)
colnames(stepsByDay) <- c("Date", "Steps")

plot(stepsByDay$Date,
     stepsByDay$Steps,
     type = "h",
     lwd=8,
     main = "Total of Steps per Day",
     xlab = "Date",
     ylab = "Frequency")

meanSteps <- mean(stepsByDay$Steps, na.rm=TRUE)
medianSteps <- median(stepsByDay$Steps, na.rm = TRUE)

```

The mean of the total number of steps per day is **`r meanSteps`**.  
The median of the total number of steps per day is **`r meanSteps`**.

## What is the average daily activity pattern?

```{r averagePattern}

stepsByInterval <- aggregate(actData$steps, by = list(actData$interval), FUN = mean, na.rm = TRUE)
colnames(stepsByInterval) <- c("interval", "avgSteps")

plot(stepsByInterval$interval,
     stepsByInterval$avgSteps,
     type = "l",
     xlab = "Reading Intervals",
     ylab = "Average Number of Steps",
     main = "Average Daily Activity Pattern")


maxPos <- which(stepsByInterval$avgSteps == max(stepsByInterval$avgSteps))
maxAvgStepsInterval <- stepsByInterval$interval[maxPos]
```

The **`r maxAvgStepsInterval`th** 5-minute interval is the interval that, on average across all the days in the dataset, contains the maximum number of steps



## Inputing missing values


The total number of missing values in the dataset is `r sum(is.na(actData$steps))`

Missing values will be filled with the rounded value of the daily average number of steps
at the corresponding 5-minute interval.

```{r}

filledData <- actData

for (i in c(1:length(filledData$steps))){
  if(is.na(filledData$steps[i])){
    nRow = i %% 288
    if (nRow == 0){
      nRow <- 288
    }
    filledData$steps[i] <- round(stepsByInterval$avgSteps[nRow])
    i <- i + 1
  }
}

stepsByDayFil <- aggregate(filledData$steps, by = list(filledData$date), FUN = sum)
colnames(stepsByDayFil) <- c("Date", "Steps")

plot(stepsByDayFil$Date,
     stepsByDayFil$Steps,
     type = "h",
     lwd=8,
     main = "Total of Steps per Day after Filling Missing Values",
     xlab = "Date",
     ylab = "Frequency")

meanStepsFil <- mean(stepsByDayFil$Steps, na.rm=TRUE)
medianStepsFil <- median(stepsByDayFil$Steps)

```
The mean of the total number of steps per day after filling missing values is **`r meanStepsFil`**.  
The median of the total number of steps per day after filling missing values is **`r medianStepsFil`**.





## Are there differences in activity patterns between weekdays and weekends?
